Follow-up Samples

### 2017-05-13 submission/manuscript.Rmd
revised methods: model building

***Model Building.*** [Comment: This entire section is pretty confusing. In the first paragraph it jumbles the treatment models and the diagnosis models]
The Random Forest [@breiman_random_2001] algorithm was used to create the three models used. The adenoma model classified normal versus adenoma, the advanced adenoma model classified normal versus advanced adenoma, and the carcinoma model classified normal versus carcinoma. The total number of individuals in the treatment models was `r length(rownames(good_metaf))` individuals.  There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the treatment carcinoma model [Table 1].  The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2]. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the carcinoma model [Table 2]. Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] which assesed the prediction of pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`), pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`).

The model included only OTU data obtained from 16S rRNA sequencing and was constructed using the R package caret (v`r paste(packageVersion("caret"))`). Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where normal and adenoma, normal and advanced adenoma, or normal and carcinoma were represented in the same proportion within both the whole data set and the 80/20 split. For each of the different splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic). The resulting model was then tested on the hold out data obtained from the 20% component. All three models had an optimized mtry of `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`. The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model.

Assessment of the most important OTUs to the model involved counting the number of times an OTU was present in the top 10% of mean decrease in accuracy (MDA) for each of the 100 different splits run. This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs. The final collated list of variables was then run through the mtry optimization again. Once the ideal mtry was found the entire sample set specific to normal versus adenoma, normal versus advanced adenoma, or normal versus carcinoma was used to create the final Random Forest model on which classifications on the `r length(rownames(good_metaf))`-person cohort was completed. For all three models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`. The only difference other than the sample set used between treatment models and normal versus diagnosis models was that only the top 10 OTUs were used to build each respective treatment model by diagnosis group to help avoid model overfitting.

>>>>

***Model Building.*** The Random Forest [@breiman_random_2001] algorithm was used to create the three models used to classify pre and post-treatment samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The total number of individuals in the pre versus post-treatment models was `r length(rownames(good_metaf))` individuals.  There were a total of `r length(rownames(filter(good_metaf, Dx_Bin == "adenoma")))` individuals in the pre versus post-treatment adenoma model, `r length(rownames(filter(good_metaf, Dx_Bin == "adv_adenoma")))` individuals in the pre versus post-treatment advanced adenoma model, and `r length(rownames(filter(good_metaf, Dx_Bin == "cancer")))` individuals in the pre versus post-treatment carcinoma model [Table 1].

Similarily, the Random Forest [@breiman_random_2001] algorithm was also used to create the three models used to classify normal versus diagnosis. The total number of individuals in the normal versus diagnosis models was `r length(rownames(metaI)) - length(rownames(good_metaf))` individuals [Table 2]. There were a total of `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus adenoma model, `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "adv Adenoma" | dx == "normal"))[, "sample"])` individuals in the normal versus advanced adenoma model, and `r length(filter(metaI, !(sample %in% c(good_metaf$initial, good_metaf$followUp)) & (Dx_Bin == "Cancer" | dx == "normal"))[, "sample"])` individuals in the normal versus carcinoma model [Table 2].

All models included only OTU data obtained from 16S rRNA sequencing and were processed and cleaned using the R package caret (v`r paste(packageVersion("caret"))`). Optimization of the mtry hyper-parameter involved making 100 different 80/20 (train/test) splits of the data where the same proportion was present within both the whole data set and the 80/20 split. For each of the different splits, 20 repeated 10-fold cross validation was performed on the 80% component to optimize the mtry hyper-parameter by maximizing the AUC (Area Under the Curve of the Receiver Operator Characteristic). The resulting model was then tested on the hold out data obtained from the 20% component. For all pre versus post-treatment models the optimized mtry was `r ifelse(as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_treat_AUC_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")` and for all normal versus diagnosis models the optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`. The hyper-parameter, mtry, defines the number of variables to investigate at each split before a new division of the data was created with the Random Forest model [@breiman_random_2001].

For each of the pre versus post-treamtent models assessment of the most important OTUs was then made by taking the top 10 OTUs by mean decrease in accuracy (MDA). These were then used to build each respective reduced OTU pre versus post-treatment model by diagnosis group to help avoid model overfitting. These reduced models were then put through the same process mentioned in the previous paragraph and were what was used for the final pre versus post-treatment models. For the normal versus diagnosis models the important OTUs were obtained by counting the number of times an OTU was present in the top 10% of MDA for each of the 100 different splits. This was then followed with filtering of this list to variables that were only present in more than 50% of these 100 runs. These corresponding reduced OTU normal versus diagnosis models were then put through the same process mentioned in the previous paragraph and were what was used for the final normal versus diagnosis models. For the pre versus post-treatment models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")` and for the normal versus diagnosis models the final optimized mtry was `r ifelse(as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(srn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] & as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"] == as.data.frame(count(crc_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], as.data.frame(count(adn_AUC_run_summary, best_mtry) %>% slice(which.max(n)))[, "best_mtry"], "not the same")`.

Each model was then applied to our `r length(rownames(good_metaf))`-person cohort [Table 1] based on diagnosis: adenoma (pre-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment adenoma (adenoma n = `r length(filter(good_metaf, Dx_Bin == "adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adenoma")[, "followUp"])`)), advanced adenoma pre-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])` ) versus post-treatment advanced adenoma (advanced adenoma n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "followUp"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma")[, "followUp"])`), and carcinoma (pre-treatment carcinoma (carinoma n = `r length(filter(good_metaf, Dx_Bin == "cancer")[, "initial"])` and disease free n = `r length(filter(good_metaf, Dx_Bin == "adv_adenoma" & Disease_Free == "n")[, "EDRN"])`) versus post-treatment carcinoma (carcinoma n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free == "n")))` and disease free n = `r length(rownames(filter(good_metaf, Dx_Bin == "cancer" & Disease_Free != "n")))`)). The application of the pre versus post-treatment models generated the probabilites that the sample was a pre-treatment sample.  The application of the normal versus diagnosis models generated the probabilities that the sample was that specific diagnosis (adenoma, advanced adenoma, or carcinoma).

### 2017-05-13 submission/manuscript.tex
revised methods: model building

\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
problem. Despite growing evidence that members of the gut microbiota can
drive tumorigenesis, little is known about what happens to the
microbiota after treatment for an adenoma or carcinoma. This study
tested the hypothesis that treatment for adenoma or carcinoma alters the
abundance of bacterial populations associated with disease to those
associated with a normal colon. We tested this hypothesis by sequencing
the 16S rRNA genes in the feces of 67 individuals before and after
treatment for adenoma (N=22), advanced adenoma (N=19), and carcinoma
(N=26).

@@

\textbf{\emph{Model Building.}} {[}Comment: This entire section is
pretty confusing. In the first paragraph it jumbles the treatment models
and the diagnosis models{]} The Random Forest {[}33{]} algorithm was
used to create the three models used. The adenoma model classified
normal versus adenoma, the advanced adenoma model classified normal
versus advanced adenoma, and the carcinoma model classified normal
versus carcinoma. The total number of individuals in the treatment
models was 67 individuals. There were a total of 22 individuals in the
treatment adenoma model, 19 individuals in the treatment advanced
adenoma model, and 26 individuals in the treatment carcinoma model
{[}Table 1{]}. The total number of individuals in the normal versus
diagnosis models was 423 individuals {[}Table 2{]}. There were a total
of 239 individuals in the adenoma model, 262 individuals in the advanced
adenoma model, and 266 individuals in the carcinoma model {[}Table 2{]}.
Each model was then applied to our 67-person cohort {[}Table 1{]} which
assesed the prediction of pre-treatment adenoma (adenoma n = 22 and
disease free n = 0) versus post-treatment adenoma (adenoma n = 0 and
disease free n = 22), pre-treatment advanced adenoma (advanced adenoma n
= 19 and disease free n = 0 ) versus post-treatment advanced adenoma
(advanced adenoma n = 0 and disease free n = 19), pre-treatment
carcinoma (carinoma n = 26 and disease free n = 0) versus post-treatment
carcinoma (carcinoma n = 1 and disease free n = 25).

The model included only OTU data obtained from 16S rRNA sequencing and
was constructed using the R package caret (v6.0.73). Optimization of the
mtry hyper-parameter involved making 100 different 80/20 (train/test)
splits of the data where normal and adenoma, normal and advanced
adenoma, or normal and carcinoma were represented in the same proportion
within both the whole data set and the 80/20 split. For each of the
different splits, 20 repeated 10-fold cross validation was performed on
the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. All three models had an optimized mtry of 2. The
hyper-parameter, mtry, defines the number of variables to investigate at
each split before a new division of the data was created with the Random
Forest model.

Assessment of the most important OTUs to the model involved counting the
number of times an OTU was present in the top 10\% of mean decrease in
accuracy (MDA) for each of the 100 different splits run. This was then
followed with filtering of this list to variables that were only present
in more than 50\% of these 100 runs. The final collated list of
variables was then run through the mtry optimization again. Once the
ideal mtry was found the entire sample set specific to normal versus
adenoma, normal versus advanced adenoma, or normal versus carcinoma was
used to create the final Random Forest model on which classifications on
the 67-person cohort was completed. For all three models the final
optimized mtry was 2. The only difference other than the sample set used
between treatment models and normal versus diagnosis models was that
only the top 10 OTUs were used to build each respective treatment model
by diagnosis group to help avoid model overfitting.

>>>>

\textbf{Background.} Colorectal cancer (CRC) is a worldwide health
problem. Despite growing evidence that members of the gut microbiota can
drive tumorigenesis, little is known about what happens to the
microbiota after treatment for an adenoma or carcinoma. This study
tested the hypothesis that treatment for adenoma or carcinoma alters the
abundance of bacterial populations associated with disease to those
associated with a normal colon. We tested this hypothesis by sequencing
the 16S rRNA genes in the feces of 67 individuals before and after
treatment for adenoma (N = 22), advanced adenoma (N = 19), and carcinoma
(N = 26).

@@

\textbf{\emph{Model Building.}} The Random Forest {[}33{]} algorithm was
used to create the three models used to classify pre and post-treatment
samples by diagnosis (adenoma, advanced adenoma, or carcinoma). The
total number of individuals in the pre versus post-treatment models was
67 individuals. There were a total of 22 individuals in the pre versus
post-treatment adenoma model, 19 individuals in the pre versus
post-treatment advanced adenoma model, and 26 individuals in the pre
versus post-treatment carcinoma model {[}Table 1{]}.

Similarily, the Random Forest {[}33{]} algorithm was also used to create
the three models used to classify normal versus diagnosis. The total
number of individuals in the normal versus diagnosis models was 423
individuals {[}Table 2{]}. There were a total of 239 individuals in the
normal versus adenoma model, 262 individuals in the normal versus
advanced adenoma model, and 266 individuals in the normal versus
carcinoma model {[}Table 2{]}.

All models included only OTU data obtained from 16S rRNA sequencing and
were processed and cleaned using the R package caret (v6.0.73).
Optimization of the mtry hyper-parameter involved making 100 different
80/20 (train/test) splits of the data where the same proportion was
present within both the whole data set and the 80/20 split. For each of
the different splits, 20 repeated 10-fold cross validation was performed
on the 80\% component to optimize the mtry hyper-parameter by maximizing
the AUC (Area Under the Curve of the Receiver Operator Characteristic).
The resulting model was then tested on the hold out data obtained from
the 20\% component. For all pre versus post-treatment models the
optimized mtry was 2 and for all normal versus diagnosis models the
optimized mtry was 2. The hyper-parameter, mtry, defines the number of
variables to investigate at each split before a new division of the data
was created with the Random Forest model {[}33{]}.

For each of the pre versus post-treamtent models assessment of the most
important OTUs was then made by taking the top 10 OTUs by mean decrease
in accuracy (MDA). These were then used to build each respective reduced
OTU pre versus post-treatment model by diagnosis group to help avoid
model overfitting. These reduced models were then put through the same
process mentioned in the previous paragraph and were what was used for
the final pre versus post-treatment models. For the normal versus
diagnosis models the important OTUs were obtained by counting the number
of times an OTU was present in the top 10\% of MDA for each of the 100
different splits. This was then followed with filtering of this list to
variables that were only present in more than 50\% of these 100 runs.
These corresponding reduced OTU normal versus diagnosis models were then
put through the same process mentioned in the previous paragraph and
were what was used for the final normal versus diagnosis models. For the
pre versus post-treatment models the final optimized mtry was 2 and for
the normal versus diagnosis models the final optimized mtry was 2.

Each model was then applied to our 67-person cohort {[}Table 1{]} based
on diagnosis: adenoma (pre-treatment adenoma (adenoma n = 22 and disease
free n = 0) versus post-treatment adenoma (adenoma n = 0 and disease
free n = 22)), advanced adenoma pre-treatment advanced adenoma (advanced
adenoma n = 19 and disease free n = 0 ) versus post-treatment advanced
adenoma (advanced adenoma n = 0 and disease free n = 19), and carcinoma
(pre-treatment carcinoma (carinoma n = 26 and disease free n = 0) versus
post-treatment carcinoma (carcinoma n = 1 and disease free n = 25)). The
application of the pre versus post-treatment models generated the
probabilites that the sample was a pre-treatment sample. The application
of the normal versus diagnosis models generated the probabilities that
the sample was that specific diagnosis (adenoma, advanced adenoma, or
carcinoma).
